---
title: "Capstone - 2021 Data"
author: "Magdalena Glotzer"
date: "2024-09-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown Document for My Capstone on 2021 Divvy Data

## Context:

Cyclistic’s finance analysts have concluded that annual members are much
more profitable than casual riders. The Director of Marketing believes
there is a solid opportunity to convert casual riders into members. The
Goal is to design marketing strategies aimed at converting casual riders
into annual members.

## Key Task:

Identify how annual members and casual riders differ in their use of the
bike-share program

## Summary of Analysis:

Analyzed member versus casual rider data for number of rides by time of
day, number of rides by day of week, number of rides per month, average
ride duration (in minutes) by day of week, average ride duration (in
minutes) by month.

## Code

### Load the necessary libraries

```{r}
library(tidyverse)  #helps wrangle data
# Use the conflicted package to manage conflicts
library(conflicted)

# Set dplyr::filter and dplyr::lag as the default choices
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

### Set Working Diretory

```{r}
setwd("~/Desktop/capstone/Divvy_Data_2021")
```

### Step 1: Collect Data

Upload Divvy datasets (csv files).

On Kaggle: 2021 data can be found here:
/kaggle/input/cyclistic-case-study-google-certificate

Downloaded the data to my drive from here: Pulled from
[here](https://divvy-tripdata.s3.amazonaws.com/index.html)

License: [here](https://divvybikes.com/data-license-agreement)

```{r}
jan <- read_csv("202101-divvy-tripdata.csv")
feb <- read_csv("202102-divvy-tripdata.csv")
mar <- read_csv("202103-divvy-tripdata.csv")
apr <- read_csv("202104-divvy-tripdata.csv")
may <- read_csv("202105-divvy-tripdata.csv")
jun <- read_csv("202106-divvy-tripdata.csv")
jul <- read_csv("202107-divvy-tripdata.csv")
aug <- read_csv("202108-divvy-tripdata.csv")
sep <- read_csv("202109-divvy-tripdata.csv")
oct <- read_csv("202110-divvy-tripdata.csv")
nov <- read_csv("202111-divvy-tripdata.csv")
dec <- read_csv("202112-divvy-tripdata.csv")
```

Check the structure of the files using spec()

```{r}
spec(jan)
spec(feb)
spec(mar)
spec(apr)
spec(may)
spec(jun)
spec(jul)
spec(aug)
spec(sep)
spec(oct)
spec(nov)
spec(dec)
```

### Step 2: Combine Datasets

```{r}
all_trips <- rbind(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec)
```

### Step 3: Clean up and add data to prepare for analysis

Inspect the new table that has been created -- the column names, the
first few lines of data, the types of data in each column, and a
statistical summary of numeric data.

```{r}
colnames(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

Add columns that list the date, hour (convert to numeric), day, month
(convert to numeric), and year of each ride. Also create a column that
specifies if a day is a weekday or weekend. Set levels for the week days
so that they will graph in the appropriate order. Going to change the
member-casual column name to usertype, due to personal preference.

```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$month <- as.numeric(all_trips$month) #Conversion
all_trips$hour <- format(as.POSIXct(all_trips$started_at), format = "%H")
all_trips$hour <- as.numeric(all_trips$hour) #Conversion
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$day_type <- ifelse(all_trips$day %in% c("Sat", "Sun"), "weekend", "weekday")
levels(all_trips$day) <- c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")

all_trips <- rename(all_trips, usertype = member_casual)
```

Add a column for "ride_length" calculation. Convert it to a numeric and
change calculation from seconds to minutes.

```{r}
# Add a "ride_length" calculation to all_trips (in minutes)
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

#Convert ride_length to minutes for easier calculation
all_trips$ride_length <- (all_trips$ride_length / 60)

summary(all_trips$ride_length)
```

Inspectin the ride_length column, there are some trip lengths that are
negative and some that are multiple days long. Delete any rides under 30
seconds long and over 6 hours long.

```{r}
all_trips_clean <- all_trips[!(all_trips$ride_length < .5 | all_trips$ride_length > 360),]
```

Remove the data where the start or end station name is NA

```{r}
all_trips_clean <- all_trips_clean %>% drop_na(start_station_name) %>% drop_na(end_station_name)
```

Create a csv file with the cleaned up data for future use

```{r}
write_csv(all_trips_clean, "full-2021-divvydata.csv")
```

###Step 4: Conduct Descriptive Analysis

```{r}
summary(all_trips_clean$ride_length)
```

Compare the ride length data for members versus casual riders

```{r}
paste('Mean ride length')
aggregate(all_trips_clean$ride_length ~ all_trips_clean$usertype, FUN = mean)
paste("Median ride length")
aggregate(all_trips_clean$ride_length ~ all_trips_clean$usertype, FUN = median)
paste("Max ride length")
aggregate(all_trips_clean$ride_length ~ all_trips_clean$usertype, FUN = max)
paste("Min ride length")
aggregate(all_trips_clean$ride_length ~ all_trips_clean$usertype, FUN = min)
```

Calculate how many rides were taken total, and how many were taken by
members versus casual riders.

```{r}
all_trips_clean %>%
summarise(number_of_rides = n())

all_trips_clean %>%
group_by(usertype)%>%
summarise(number_of_rides = n())%>%
ggplot(aes(x=usertype, y = number_of_rides, fill = usertype))+
geom_col(position = "dodge") + geom_text(aes(label = number_of_rides)) + 
labs(caption = "4,545,092 as total number of rides ")
```

Number of rides by bike type (with detail of casual and member counts)

```{r}
all_trips_clean %>%
group_by(usertype, rideable_type)%>%
summarise(number_of_rides = n())%>%
ggplot(aes(x=rideable_type, y = number_of_rides, fill = usertype))+
geom_col(position="stack")
```

Calculate average ride time for all riders. Visualize by usertype

```{r}
all_trips_clean %>%
summarise(average_duration = mean(ride_length))

all_trips_clean %>%
group_by(usertype)%>%
summarise(average_duration = mean(ride_length))%>%
ggplot(aes(x=usertype, y = average_duration, fill = usertype))+
geom_col(position = "dodge")+
geom_label(aes(x = usertype, label=average_duration)) + 
labs(caption = "Average ride time overall is 19.14989 minutes")
```

Create dataframe with average ride time per day of week for members vs
casual users. This doesn't sort the data correctly, but it's a good
starting point.

```{r}
aggregate(all_trips_clean$ride_length ~ all_trips_clean$usertype + 
          all_trips_clean$day_of_week, FUN = mean)
```

Create a dataframe that calculates 1. average ride time and 2. number of
rides, per day of week for members and casual users. This dataframe will
be sorted and presented Sunday to Saturday.

```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>%
  arrange(weekday, usertype)
```

Create a dataframe that calculates 1. average ridetime and 2. number of
rides, per month for members and casual users. This dataframe will be
sorted and printed January through December.

```{r}
all_trips_clean %>% 
  mutate(month = month(started_at, label = TRUE)) %>%
  group_by(usertype, month) %>% 
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>%
  arrange(month, usertype)
```

Create a visualization showing average number of rides per day of the
week. Separate by usertype.

```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = usertype)) +
  facet_wrap(~usertype) +
  geom_col(position = "dodge")
```

Create a visualization showing average ride duration per day of the
week. Separate by usertype.

```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = usertype)) + 
  facet_wrap(~usertype)+
  geom_col(position = "dodge")
```

Create a visualization showing number of ride by month. Note that June
data is incomplete as the first datapoint is from 2013-06-27.

```{r}
all_trips_clean %>% 
  mutate(month = month(started_at, label = TRUE)) %>% 
  group_by(usertype, month) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, month)  %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")
```

Create a visualization showing average ride duration by month. Note that
June data is incomplete as the first datapoint is from 2013-06-27.

```{r}
all_trips_clean %>% 
  mutate(month = month(started_at, label = TRUE)) %>% 
  group_by(usertype, month) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, month)  %>% 
  ggplot(aes(x = month, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge")
```

Create a visualization that shows the number of rides taken over the
course of the day.

```{r}
all_trips_clean %>% 
  mutate(time_of_day = as.numeric(hour)) %>% 
  group_by(usertype, time_of_day) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(usertype, time_of_day)  %>% 
  ggplot() +
  geom_line(aes(x = time_of_day, y = number_of_rides, color = usertype))
```

Create a visualization that shows the duration of rides taken over the
course of the day.

```{r}
all_trips_clean %>% 
  mutate(time_of_day = as.numeric(hour)) %>% 
  group_by(usertype, time_of_day) %>% 
  summarise(average_duration = mean(ride_length)) %>% 
  arrange(usertype, time_of_day)  %>% 
  ggplot() +
  geom_line(aes(x = time_of_day, y = average_duration, color = usertype))
```

Looking at start and end points: Given the question we are trying to
solve for this capstone, it would be useful to know which stations are
most popular for casual users versus members.

Most popular start station for members:

```{r}
all_trips_clean %>%
   filter(usertype == "member") %>% 
group_by(start_station_name) %>% 
summarise(number_of_rides = n()) %>% 
arrange(desc(number_of_rides)) %>%
slice(1:5)
```

Most popular start station for casual users:

```{r}
all_trips_clean %>%
   filter(usertype == "casual") %>% 
group_by(start_station_name) %>% 
summarise(number_of_rides = n()) %>% 
arrange(desc(number_of_rides)) %>%
slice(1:5)
```

Most popular end station for members:

```{r}
all_trips_clean %>%
   filter(usertype == "member") %>% 
group_by(end_station_name) %>% 
summarise(number_of_rides = n()) %>% 
arrange(desc(number_of_rides)) %>%
slice(1:5)
```

Most popular end station for casual users:

```{r}
all_trips_clean %>%
  filter(usertype == "casual") %>% 
group_by(end_station_name) %>% 
summarise(number_of_rides = n()) %>% 
arrange(desc(number_of_rides)) %>%
slice(1:5)
```

Most popular start station overall:

```{r}
all_trips_clean %>%
group_by(start_station_name) %>% 
summarise(number_of_rides = n()) %>% 
arrange(desc(number_of_rides)) %>%
slice(1:5)
```

Most popular end station overall:

```{r}
all_trips_clean %>%
group_by(end_station_name) %>% 
summarise(number_of_rides = n()) %>% 
arrange(desc(number_of_rides)) %>%
slice(1:5)
```

The data above shows that the most popular start and end stations for
casual riders are at tourist destinations: Navy Pier (Streeter Dr &
Grand Ave), Millenium Park, an intersection by Water Tower Place and the
Drake Hotel (Michigan Ave & Oak St), Theatre on the Lake, and the Shedd
Aquarium.
